### 2020-06-08 submission/manuscript.Rmd
Fix code chunk labels, add code to figure legends to indicate N for analyses

```{r, cfu}
cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
mean_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "mean_cfu", names_to = "Source") %>% arrange(mean_cfu)
min_mean_d1 <- mean_d1 %>%  select(mean_cfu) %>% min
max_mean_d1 <- mean_d1 %>% select(mean_cfu) %>% max

@@

```{r, alpha}
richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities vary across mouse colony sources.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r, adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

adonis_stats_dn101 <- read_tsv("../data/process/adonis_dn1-1.tsv")
dn1_rsq_sum <- adonis_stats_dn101 %>% filter(day == -1) %>% select(r_sq) %>% sum
d0_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 0) %>% select(r_sq) %>% sum
d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor communities were relatively stable between the 2 experiments. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.
```{r, source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

```{r, source_clindamycin}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()

@@

```{r, classification}
models_vs_random <- read_tsv("../data/process/classification_to_random.tsv")
max_p_vs_random <- models_vs_random %>% select(p.value) %>% max
median_otu_day0_AUROC <- models_vs_random %>% filter(model_name == "otu_day0") %>% pull(median)

@@

## Figures
```{r, sample_size}
#figure out in functions.R and export results as a tsv to read in
#analysis_n <- read_tsv("../data/process/____.tsv")

```
\includegraphics{figure_1.pdf}
**Figure 1. Clindamycin is sufficient to promote *C. difficile* colonization regardless of colony source, but clearance time varies across sources of C57BL/6 mice.** A. Setup of the experimental timeline. Mice for the experiments were obtained from 6 different sources: the Schloss (N =8) and Young lab (N = 9) colonies at the University of Michigan, the Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic Biosciences (N = 8), and Envigo (N = 8). All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C. difficile* 630 spores on day 0. Mice were weighed and feces was collected daily through the end of the experiment (9 days post-infection). Note: 3 mice died during course of experiment. 1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse between 1- and 3-days post-infection. B. *C. difficile* CFU/gram stool measured over time (N = 20-49 mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the serial dilutions. C. Mouse weight change measured in grams over time (N = 45-49 mice per timepoint), all mice were normalized to the weight recorded 1 day before infection. For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001). Lines represent the mean for each source and circles represent individual mice. D. *C. difficile* CFU status on Day 7 within each mouse colony source. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU). Schloss (N = 6), Young (N = 9), Jackson (N = 5), Charles River (N = 8), Taconic (N = 5), Envigo (N = 7). Alternative: N = 5-9 mice per group.

@@

\newpage
\includegraphics{figure_5.pdf}
**Figure 5. Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B. Boxplots of the top 10 most significant (adjusted P value < 0.05) OTUs with relative abundances that changed post clindamycin treatment. C-D. Boxplots of the top 10 most significant families with relative abundances that changed post clindamycin treatment. Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = 31). Tests were performed at the OTU and family level with Benjamini-Hochberg correction for testing all identified OTUs and families. See Table S10-11 for complete list of OTUs and families significantly impacted by clindamycin treatment. The grey vertical line indicates the limit of detection for A-D.

@@

\newpage
\includegraphics{figure_S2.pdf}
**Figure S2. Only bacterial communities from University of Michigan mice significantly vary between experiments.** A-F. PCoA of Theta YC distances for the baseline fecal bacterial communities within each source of mice. Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates. PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation. Experiment number and cage only significantly explained observed variation for mice from the Schloss (add r2 and P values) and Young (add r2 and P values) lab colonies (Table S5).

\newpage
\includegraphics{figure_S3.pdf}
**Figure S3. *C. difficile* colonization status 7 days post-infection varies across mice.** A. Bar graph visualizations of overall day 7 classification outcomes used to build logistic regression models. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection. B-D. Classification input data used for models based on day -1 (B), day 0 (C), and day 1 (D) community relative abundance data. Only mice with day 7 *C. difficile* status and community relative abundance data for each respective timepoint were used as input data for the classification models.

@@

**Movie S1. Large shifts in bacterial community structure occurred after clindamycin and *C. difficile* infection.** PCoA of Theta YC distances animated from 0 through 9 days post-infection. Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint. See Table S5 for PERMANOVA results).

>>>>

```{r cfu}
cfu_stats_all <- read_tsv("../data/process/cfu_stats_all_days.tsv")
padj_d1_cfu <- cfu_stats_all %>% filter(day == 1) %>% pull(p.value.adj) 
mean_d1 <- cfu_stats_all %>% filter(day == 1) %>% select(-day, -statistic, -p.value, -parameter, -method, -p.value.adj) %>% pivot_longer(everything(), values_to = "mean_cfu", names_to = "Source") %>% arrange(mean_cfu)
min_mean_d1 <- mean_d1 %>%  select(mean_cfu) %>% min
max_mean_d1 <- mean_d1 %>% select(mean_cfu) %>% max

@@

```{r alpha}
richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities vary across mouse colony sources.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

adonis_stats_dn101 <- read_tsv("../data/process/adonis_dn1-1.tsv")
dn1_rsq_sum <- adonis_stats_dn101 %>% filter(day == -1) %>% select(r_sq) %>% sum
d0_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 0) %>% select(r_sq) %>% sum
d1_rsq_sum <-  adonis_stats_dn101 %>% filter(day == 1) %>% select(r_sq) %>% sum
p_adonis_intial <- adonis_stats_dn101 %>% select(p) %>% max 
p_adonis_min <- adonis_stats_dn101 %>% select(p) %>% min #All p values are same

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor communities were relatively stable between the 2 experiments. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.
```{r source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

```{r clindamycin_taxa}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()

@@

```{r classification}
models_vs_random <- read_tsv("../data/process/classification_to_random.tsv")
max_p_vs_random <- models_vs_random %>% select(p.value) %>% max
median_otu_day0_AUROC <- models_vs_random %>% filter(model_name == "otu_day0") %>% pull(median)

@@

## Figures
```{r sample_size}
#figure out in functions.R and export results as a tsv to read in
analysis_n <- read_tsv("../data/process/processed_metadata.tsv")

#Initial N for each source from the 2 experiments combined:
schloss_i <- analysis_n %>% filter(day == -1 & vendor == "Schloss") %>% nrow()
young_i <- analysis_n %>% filter(day == -1 & vendor == "Young") %>% nrow()
jackson_i <- analysis_n %>% filter(day == -1 & vendor == "Jackson") %>% nrow()
cr_i <- analysis_n %>% filter(day == -1 & vendor == "Charles River") %>% nrow()
taconic_i <- analysis_n %>% filter(day == -1 & vendor == "Taconic") %>% nrow()
envigo_i <- analysis_n %>% filter(day == -1 & vendor == "Envigo") %>% nrow()

#Range in N for C. difficile CFU quantification:
cfu_n_by_day <- analysis_n %>% 
  filter(!is.na(cfu)) %>% #remove rows with NAs
  group_by(day) %>% count() %>% arrange(n) %>% ungroup()
min_cfu_n <- cfu_n_by_day %>% select(n) %>% min
max_cfu_n <- cfu_n_by_day %>% select(n) %>% max

#Range in N for mouse weight change data over the course of the experiment:
weight_n_by_day <- analysis_n %>% 
  filter(!is.na(weight)) %>% #remove rows with NAs
  group_by(day) %>% count() %>% arrange(n) %>% ungroup()
min_weight_n <- weight_n_by_day %>% select(n) %>% min
max_weight_n <- weight_n_by_day %>% select(n) %>% max

#C. difficile day 7 colonization status N range across sources:
day7_status_n <- analysis_n %>% 
  filter(!is.na(clearance_status_d7)) %>% #remove rows with NAs
  filter(day == 7) %>% 
  group_by(vendor) %>% count() %>% arrange(n) %>% ungroup
day7_status_n_min <- day7_status_n %>% select(n) %>% min
day7_status_n_max <-day7_status_n %>% select(n) %>% max

#N for mice with sequencing data on both day -1 and day 0 (used to determine taxa that were altered by clindamycin treatment)
n_paired_for_clind <- read_tsv("../data/process/vendors.diversity.summary") %>% #read in sequencing data & join to metadata
  filter(method == "ave") %>% 
  select(group, sobs, shannon, invsimpson, coverage) %>% 
  inner_join(metadata, by = c("group" = "id")) %>% 
  filter(day == -1 | day == 0) %>% 
  filter(duplicated(mouse_id)) %>% 
  nrow()
```
\includegraphics{figure_1.pdf}
**Figure 1. Clindamycin is sufficient to promote *C. difficile* colonization regardless of colony source, but clearance time varies across sources of C57BL/6 mice.** A. Setup of the experimental timeline. Mice for the experiments were obtained from 6 different sources: the Schloss (N = `r schloss_i`) and Young lab (N = `r young_i`) colonies at the University of Michigan, the Jackson Laboratory (N = `r jackson_i`), Charles River Laboratory (N = `r cr_i`), Taconic Biosciences (N = `r taconic_i`), and Envigo (N = `r envigo_i`). All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C. difficile* 630 spores on day 0. Mice were weighed and feces was collected daily through the end of the experiment (9 days post-infection). Note: 3 mice died during course of experiment. 1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse between 1- and 3-days post-infection. B. *C. difficile* CFU/gram stool measured over time (N = `r min_cfu_n`-`r max_cfu_n` mice per timepoint) via serial dilutions. The black line represents the limit of detection for the first serial dilution. CFU quantification data was not available for each mouse due to early deaths, stool sampling difficulties, and not plating all of the seriral dilutions. C. Mouse weight change measured in grams over time (N = `r min_weight_n`-`r max_weight_n` mice per timepoint), all mice were normalized to the weight recorded 1 day before infection. For B-C: timepoints where differences across sources of mice were statistically significant by Kruskal-Wallis test with Benjamini-Hochberg correction for testing across multiple days (Table S1 and Table S2) are reflected by the asterisk(s) above each timepoint (\*\, P < 0.05, \*\*\, P < 0.01, \*\*\*\, P < 0.001). Lines represent the mean for each source and circles represent individual mice. D. *C. difficile* CFU status on Day 7 within each mouse colony source. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU). N = `r day7_status_n_min`-`r day7_status_n_max` mice per group.

@@

\newpage
\includegraphics{figure_5.pdf}
**Figure 5. Clindamycin treatment has the same effects on a subset of taxa regardless of colony source.** A-B. Boxplots of the top 10 most significant (adjusted *P* value < 0.05) OTUs with relative abundances that changed post clindamycin treatment. C-D. Boxplots of the top 10 most significant families with relative abundances that changed post clindamycin treatment. Data were analyzed by Wilcoxon signed rank test limited to mice that had paired sequence data for day -1 and 0 (N = `r n_paired_for_clind`). Tests were performed at the OTU and family levels with Benjamini-Hochberg correction for testing all identified OTUs and families. See Table S10-11 for complete list of OTUs and families significantly impacted by clindamycin treatment. The grey vertical line indicates the limit of detection for A-D.

@@

\newpage
\includegraphics{figure_S2.pdf}
**Figure S2. Only bacterial communities from University of Michigan mice significantly vary between experiments.** A-F. PCoA of Theta YC distances for the baseline fecal bacterial communities within each source of mice. Each symbol represents a stool sample from an individual mouse with color corresponding to experiment and shape representing cage mates. PERMANOVA was performed within each group to examine the contributions of experiment and cage to observed variation. Experiment number and cage only significantly explained observed variation for mice from the Schloss (combined R^2^ = `r adonis_schloss_rsq_sum`; *P* $\le$ `r p_adonis_schloss_max`) and Young (combined R^2^ = `r adonis_young_rsq_sum`; *P* $\le$ `r p_adonis_young_max`) lab colonies (Table S7).

\newpage
\includegraphics{figure_S3.pdf}
**Figure S3. *C. difficile* colonization status 7 days post-infection varies across mice.** A. Bar graph visualizations of overall day 7 C. difficile colonization status that were used as classification outcomes to build logistic regression models. Mice were classified as colonized or cleared (not detectable at the limit of detection of 100 CFU) based on CFU g/stool data from 7 days post-infection. B-D. Classification input data used for models based on day -1 (B), day 0 (C), and day 1 (D) community relative abundance data. Only mice with day 7 *C. difficile* status and community relative abundance data for each respective timepoint were used as input data for the classification models.

@@

**Movie S1. Large shifts in bacterial community structure occurred after clindamycin and *C. difficile* infection.** PCoA of Theta YC distances animated from 0 through 9 days post-infection. PERMANOVA analysis indicated colony source was the variable that explained the most observed variation across fecal communities (source R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment. Transparency of the circle corresponds to the day of the experiment, each circle represents a sample from an individual mouse at a specific timepoint. See Table S5 for PERMANOVA results).
